UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-05 ms/main.tex
add some text on FEV1 and ciations

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations \sgcomment{point to box plots again?}.

Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan \sgcomment{if test is not about those born in Japan, needs to be clarified!} have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (see figure~\ref{fig:supp_fev_ridgeplots} for distributions). For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.

Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups. This is especially useful when population labels based on self-reported identity or country of birth are unavailable or correlate poorly with genetic ancestry.

>>>>

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations (figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m}).

\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage.}
Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. Baseline FEV1 levels are determined via equations that include ancestry or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}; we are not aware of any results comparing populations within Asia. We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines (the largest populations in the UKBB by country of birth) as well as the self-identified Chinese population (mutually exclusive groups, shown in figure~\ref{fig:supp_fev_test_pops}). After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$). By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}). For comparison, the adjusted difference between self-identified African and British individuals in the UKBB is 0.762 (p-value $2.2\times 10^{-16}$).

%For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.

%Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups. This is especially useful when population labels based on self-reported identity or country of birth are unavailable or correlate poorly with genetic ancestry.
